---
title: "Analyzing Factors Related to Voter Preferences in the United States"
author: Ravnit Lotay
date: "today"
date-format: long
abstract: "This study investigates the factors influencing voter support in the 2022 US presidential election, utilizing logistic regression on CCES 2022 data. The results of the analysis will be crucial in understanding why individuals or demographics vote for a certain individual or party,"
format: pdf
toc: true
number-sections: true
bibliography: references.bib
nocite: |
  @rstanarm, @tidyverse, @haven, @dplyr, @dataverse, @readr, @here, @gridExtra, @kableExtra, @ggplot2, @modelsummary, @knitr
---

```{r setup, include=FALSE}
library(tidyverse)
library(rstanarm)
library(kableExtra)
library(ggplot2)
library(modelsummary)
library(readr)
library(here)
library(gridExtra)
library(knitr)

cces2022 = read_csv(here::here("data/analysis_data/cces_analysis_data.csv"))
```


\newpage 


# Introduction

The goal of the Cooperative Election Study (CES) is to investigate how Americans vote and perceive their electoral experiences, how they hold their representatives accountable in elections, and how their behavior and experiences vary. The study used a sample of 60,000, allowing it to account for variations across different legislative districts.

It is important to note that the 2022 CES data has adjusted the input data via a weightening process to ensure samples are valid and representative.

# Data {#sec-data}

The Dataset contains 60,000 cases in which individuals responsed to survey questions including, but not limited to: voting prefrence, age, gender and highest level of education. The mentioned factors will be used to draw inferences about voter preferences via a logistic regression model.

## Data Visualizations

Within the CES data, there are relations between the different variables. Below are two figures showing how gender, age, and education relate to voter preference.

```{r}
#| label: fig-fig1
#| fig-cap:  Distribution of Presidential Preferences by gender and education
#| echo: false
#| warning: false
#| message: false

cces2022 |>
  ggplot(aes(x = education, fill = voted_for)) +
  stat_count(position = "dodge") +
  facet_wrap(facets = vars(gender)) +
  theme_minimal() +
  labs(
    x = "Highest education",
    y = "Number of respondents",
    fill = "Voted for"
  ) +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```
```{r}
#| label: fig-fig2
#| fig-cap: Distribution of Presidential Preferences by age and education
#| echo: false
#| warning: false
#| message: false

cces2022 |>
  mutate(age_group = case_when(
    age <= 24 ~ "18-24",
    age >= 25 & age <= 34 ~ "25-34",
    age >= 35 & age <= 44 ~ "35-44",
    age >= 45 & age <= 54 ~ "45-54",
    age >= 55 & age <= 64 ~ "55-64",
    age >= 65 ~ "65+"
  )) |>
  ggplot(aes(x = age_group, fill = voted_for)) +
  geom_bar(position = "dodge") +
  facet_wrap(facets = vars(gender)) +
  theme_minimal() +
  labs(
    x = "Age Group",
    y = "Number of Respondents",
    fill = "Voted For"
  ) +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```

# Model {#sec-model}

To assess the roles of gender, education, and age on voting preferences, a logistic regression model is used.

## Model Creation

To create the model, let $y_i$ represent the respondent's political preference, defined as 1 for Trump and 0 for Biden. Gender is denoted by $g_i$, and age by $a_i$.

The logistic regression model can be formed as follows:


\begin{align*} 
y_i|\pi_i &\sim \text{Bernoulli}(\pi_i) \\
\text{logit}(\pi_i) & = \beta_0 + \beta_1 \times g_i + \beta_2 \times e_i \\
\beta_0 &\sim \text{Normal}(0, 2.5)\\
\beta_1 &\sim \text{Normal}(0, 2.5)\\
\beta_2 &\sim \text{Normal}(0, 2.5)
\end{align*}


Where:
- $y_i$ is the political preference of the ith respondent.
- $\pi_i$ is the probability of supporting Trump for the ith respondent.
- $g_i$ represents the gender of the ith respondent.
- $e_i$ represents the education level of the ith respondent.
- $\beta_0$, $\beta_1$, and $\beta_2$ are the parameters of the model, representing the intercept, the effect of gender, and the effect of education on the likelyhood of supporting Trump, respectively.


# Results {#sec-result}

The logistic regression model provides invaluable information related to voter preferences.


## Results Table
```{r fig-fig3, echo=FALSE, warning=FALSE}
#| label: tbl-table1
#| tbl-cap: Results of logistic regression on the voted for Trump

voter_preferences <- readRDS(here::here("model/2022_cces_voter_model.rds"))

modelsummary(
  list(
    "Support Trump" = voter_preferences
   ),
   statistic = "mad"
)

```

Below, the results of the logistic regression model are shown in a statistical form.

In the next section, the results will be interpreted in a simpler manner.

# Discussion {#sec-discussion}

## Model Interpreted Results

The summary of the results, as shown above, indicate the following:

1. Females and those identifying as "Non-binary" are more strongly against voting for Trump when compared to the reference group.
   
2. Those identifying as "Other" are more likely to vote for Trump than the reference group.
   
3. Those that have completed a 4-year college or university program are much less likely to vote for Trump than those with no highschool education or those with a highschool diploma.
   
The coefficient for woman, is seen as -0.427, while the coefficient for Non-binary is -2.274. This, in essence, implies that these groups are much more opposed to voting for Trump than our baseline reference group of males. 

As for age, due to the positive coefficient, it is evident that the older an individual is, the more likely they are to vote for Trump.

It is important to note that since these values are based on a reference group (males for the gender attribute), a positive coefficient does not inherently mean that the majority of the group is voting for Trump. To further prove this statement, it is clear that despite many coefficients being positive in the above table, Donald Trump was not elected in the last presidential election. 

## Weaknesses

Considering the data collected and analyzed, there are several weaknesses that could be addressed. First, there are relatively few Non-binary and Other values in the gender column, and as such they could be grouped together or otherwise adjusted for. This is necessary as logistic regression is typically influenced greatly by outliers. Second, additional parameters, such as race, could be used to improve the quality of the model.

\newpage

# References

\nocite{rstanarm, tidyverse, haven, dplyr, dataverse, readr, here, gridExtra, kableExtra, ggplot2, modelsummary, knitr}